---
title: "Early detection and risk stratification of heart disease"
author: "Luca Signorini, Samuel Frei, Julia Studer"
date: "15.11.2025"
output:
  html_document:
    highlight: tango   # for line numbering to work
---




# Introduction

Motivation, research question, context

Use case: A regional hospital wants to improve its preventive care and identify patients with an increased risk of heart disease at an early stage.
Until now, risk assessment has mainly been carried out by doctors on the basis of individual factors (age, BMI, blood pressure). However, the hospital wants to understand:
Which factors have the greatest influence on the risk of heart disease
How these factors reinforce or weaken each other
Whether modern machine learning models provide better predictions than traditional statistics
At what thresholds the risk increases significantly
How patients can be given targeted preventive care
To this end, the hospital has commissioned our project team to analyze the Person Health Indicator dataset.

The overarching question of the project
“How well can the risk of heart disease be predicted using personal health indicators, and which factors have the strongest influence?”
We answer this question in two dimensions:
I. Risk prediction (predictive task)
Can a model reliably identify patients at increased risk?
→ For this: GLM Binomial, SVM, Neural Network
II. Risk explanation (interpretation task)
Which factors drive the risk, and how do these effects unfold?
→ For this: Linear Model, Poisson GLM, GAM 

What the hospital ultimately receives
A clear, clinically useful answer:
Early warning model: Who is at risk?
Top 5 main risk factors: What drives the risk?
Threshold values: When does a factor become critical?
Recommendations: Where is patient intervention worthwhile (exercise, BMI, sleep)?
Model comparison: Which model is the most reliable?

<br>
--------------------------------------------------------------------------------

<br>

# 1. Data Preparation

**Dataset:** heart_2022_no_nans.csv
<br>
**Source:** https://www.kaggle.com/datasets/kamilpytlak/personal-key-indicators-of-heart-disease?resource=download&select=2022

After loading the dataset, we reduce both the number of variables and observations to stay within the task limits. To address the class imbalance in the target variable HadHeartAttack (more “No” than “Yes”), we apply undersampling to the majority class, achieving a 1:4 ratio between “Yes” and “No” cases.

<br>
<details>
<summary><b>Show code: Prerequisites</b></summary>
```{r prerequisites, cache=TRUE}
# Prerequisites
#install.packages("dplyr")
library(dplyr)
#install.packages("skimr")
library(skimr)
```
</details>

<details>
<summary><b>Show code: Data import and summary</b></summary>
```{r loading-data, cache=TRUE}
# Loading data
#data <- read.csv("heart_2022_no_nans.csv")
data <- read.csv(file = "/Users/lucasignorini/Documents/MSc_ids/Sem_3/ML1/Gruppenarbeit/Data/archive/2022/heart_2022_no_nans.csv")

skim(data)
```
</details>

```{r recode-outcome, cache=TRUE}
# Ensure outcome is a proper binary factor
data$HadHeartAttack <- factor(data$HadHeartAttack,
                              levels = c("No", "Yes"))

# Quick check
str(data$HadHeartAttack)
table(data$HadHeartAttack)
```

<details>

```{r convert-types, cache=TRUE}
data <- data %>%
  mutate(
    HadHeartAttack     = factor(HadHeartAttack, levels = c("No", "Yes")),
    AgeCategory        = factor(AgeCategory),
    Sex                = factor(Sex),
    SmokerStatus       = factor(SmokerStatus),
    HadDiabetes        = factor(HadDiabetes),
    PhysicalActivities = factor(PhysicalActivities),
    GeneralHealth      = factor(GeneralHealth),
    DifficultyWalking  = factor(DifficultyWalking),
    HadKidneyDisease   = factor(HadKidneyDisease),
    CovidPos           = factor(CovidPos),
    BMI                = as.numeric(BMI),
    SleepHours         = as.numeric(SleepHours)
  )

str(data$AgeCategory)
```

<summary><b>Show code: Select predictors</b></summary>
```{r select-predictors, cache=TRUE}
# Keep only 11 selected predictors = 12 variables
df <- data.frame(data$HadHeartAttack,
                 data$AgeCategory,
                 data$Sex,
                 data$BMI,
                 data$SmokerStatus,
                 data$HadDiabetes,
                 data$PhysicalActivities,
                 data$GeneralHealth,
                 data$SleepHours,
                 data$DifficultyWalking,
                 data$HadKidneyDisease,
                 data$CovidPos)
```
</details>

<details>
<summary><b>Show code: Handling class imbalance</b></summary>
```{r handling-class-imbalance, cache=TRUE}

set.seed(123)


minority <- data %>% filter(HadHeartAttack == "Yes")
majority <- data %>% filter(HadHeartAttack == "No")


ratio <- 4   # 4:1 majority : minority


majority_under <- majority %>% sample_n(nrow(minority) * ratio)

#combine undersampled majority with minority
data <- bind_rows(minority, majority_under)

#check new class distribution
table(data$HadHeartAttack)

#sumary of reduced dataset
skim(data)
```

</details>

```{r split, cache=TRUE}
set.seed(123)
library(caret)

train_index <- createDataPartition(data$HadHeartAttack, p = 0.7, list = FALSE)

train <- data[train_index, ]
test  <- data[-train_index, ]
```

--------------------------------------------------------------------------------

<br>

# 2. Data Visualization

Here the visualization of our dataset is described.

<br>

--------------------------------------------------------------------------------

<br>

# 3. Methods 
### Data Preparation
**Handling Class Imbalance in the Target Variable**

In the original dataset, the outcome variable HadHeartAttack is highly imbalanced, with only a small fraction of participants reporting a past heart attack. This imbalance poses a major challenge for predictive models: Without correction, most models learn to simply predict “No heart attack” for all cases. While this can still produce a high overall accuracy, such models fail exactly where the hospital needs them most — identifying patients at increased cardiovascular risk.

Due to the large sample size of the provided dataset, we were required to use the undersampling method to achieve a 4:1 ratio of our target variable - 'HadHeartAttack' - as well as to meet the dataset size prerequisites. From an initial sample size of over 200'000, we reduced it to about 67'000, of which 53'740 did not have a heart attack and 13'435 did.



Description of the models 

## 3.1. Risk prediction: Can a model reliably identify patients at increased risk?

### Generalized linear model Binomial

Samuel

```{r glm-binomial, cache=TRUE}
glm_bin <- glm(
  HadHeartAttack ~ AgeCategory + Sex + BMI + SmokerStatus +
    HadDiabetes + PhysicalActivities + GeneralHealth +
    SleepHours + DifficultyWalking + HadKidneyDisease + CovidPos,
  data = train,
  family = binomial
)

summary(glm_bin)
```

```{r glm-roc, cache=TRUE}
library(pROC)

prob_glm <- predict(glm_bin, newdata = test, type = "response")

roc_glm <- roc(test$HadHeartAttack, prob_glm)

plot(roc_glm, main = "ROC Curve — GLM Binomial")
auc(roc_glm)
```

```{r glm-bmi-curve, cache=TRUE}
bmi_grid <- data.frame(
  BMI                = seq(15, 50, length.out = 150),
  AgeCategory        = levels(train$AgeCategory)[1],
  Sex                = levels(train$Sex)[1],
  SmokerStatus       = levels(train$SmokerStatus)[1],
  HadDiabetes        = levels(train$HadDiabetes)[1],
  PhysicalActivities = levels(train$PhysicalActivities)[1],
  GeneralHealth      = levels(train$GeneralHealth)[1],
  SleepHours         = 7,
  DifficultyWalking  = levels(train$DifficultyWalking)[1],
  HadKidneyDisease   = levels(train$HadKidneyDisease)[1],
  CovidPos           = levels(train$CovidPos)[1]
)

bmi_grid$prob <- predict(glm_bin, newdata = bmi_grid, type = "response")

plot(
  bmi_grid$BMI, bmi_grid$prob, type = "l", lwd = 2,
  xlab = "BMI", ylab = "Predicted probability",
  main = "Predicted Heart Attack Risk vs BMI (GLM)"
)
```



### Support Vector machines

Luca

### Neural Network 

Julia 

## 3.2. Risk explanation: Which factors drive the risk, and how do these effects unfold?

### Exploratory Linear Models Among Predictors

To better understand how the health indicators relate to one another — independently of the heart attack outcome — we fitted simple linear models using BMI and SleepHours as continuous response variables. These exploratory models help reveal clusters of risk factors and provide context for the effects observed in the GLM and GAM models.
<br>

#### Linear Model 1: Predicting BMI

We did not log-transform the BMI do it is a continous variable because it is strictly positive, just slightly right-skewed and most importantly the interpretation of the Model-Output remains intuitiv.
<br>

<details>
<summary><b>Show code: linear model predicting BMI</b></summary>
```{r lm_bim, cache=TRUE}
### ======================================================
### ORDER FACTORS IN A LOGICAL WAY
### ======================================================

# Age (chronological)
train$AgeCategory <- factor(train$AgeCategory, levels = c(
  "Age 18 to 24",
  "Age 25 to 29",
  "Age 30 to 34",
  "Age 35 to 39",
  "Age 40 to 44",
  "Age 45 to 49",
  "Age 50 to 54",
  "Age 55 to 59",
  "Age 60 to 64",
  "Age 65 to 69",
  "Age 70 to 74",
  "Age 75 to 79",
  "Age 80 or older"
))

# Ordered health categories
train$GeneralHealth <- factor(train$GeneralHealth,
                              levels = c("Excellent", "Very good", "Good", "Fair", "Poor"))

# Physical activity (nominal)
train$PhysicalActivities <- factor(train$PhysicalActivities,
                                   levels = c("Yes", "No"))

# Smoking (nominal)
train$SmokerStatus <- factor(train$SmokerStatus,
                             levels = c("Never smoked",
                                        "Former smoker",
                                        "Current smoker - now smokes some days",
                                        "Current smoker - now smokes every day"))

# Sex
train$Sex <- factor(train$Sex, levels = c("Female", "Male"))


### ======================================================
### FIT THE MODEL WITH HEALTHY BASELINE
### ======================================================

lm_bmi_healthy <- lm(
  BMI ~ AgeCategory + Sex + PhysicalActivities + SleepHours +
        GeneralHealth + SmokerStatus,
  data = train
)

summary(lm_bmi_healthy)
```
</details>
<br>

Using a healthy reference profile (young, female, active, excellent health, never smoked), the model reveals clear associations with BMI. BMI rises steadily across adulthood and peaks around ages 45–55, consistent with epidemiological patterns, before declining in the oldest age groups. Physical inactivity is associated with a substantial BMI increase of 1.7 units. Worse self-rated general health shows a strong monotonic association with higher BMI, with individuals reporting “Fair” or “Poor” health exhibiting more than 4 BMI units above the healthy baseline. Current smokers have significantly lower BMI than never-smokers, a known effect of nicotine. Although the model explains around 11% of BMI variation—typical for behavioral health data—it provides meaningful insight into how lifestyle and health characteristics cluster in ways relevant for cardiovascular risk.

<details>
<summary><b>Show detailed interpretation: linear model predicting BMI</b></summary>
<br>
**This linear model estimates how demographic, lifestyle, and health variables are associated with BMI, using a healthy individual as the reference category:**

- Age 18–24
- Female
- Physically active (PhysicalActivities = Yes)
- Excellent general health
- Never smoked

The intercept (23.84) is therefore the estimated BMI for this baseline individual.

<br>
**1. Age effects:** *BMI increases steadily with age until ~55–59, then declines in very old age*

Relative to the 18–24 baseline:

- Ages 25–29: +2.00 BMI
- Ages 40–44: +3.77 BMI
- Ages 45–49: +4.05 BMI
- Ages 50–54: +3.98 BMI
- Ages 55–59: +3.39 BMI

These are all highly significant (p < 2e-16).

This pattern reflects a typical adult BMI trajectory:
BMI rises through early and middle adulthood, stabilizes, then decreases at the oldest ages.

Age 80+ shows –0.47 BMI (p = 0.0049) versus young adults → older adults often experience muscle loss and weight reduction.

<br>
**2. Sex:** *Males have slightly higher BMI*

SexMale = +0.27 BMI units (p ≈ 2.6e-06)

This indicates a modest but significant difference, consistent with population norms.

<br>
**3. Physical activity:** *Inactive individuals have substantially higher BMI*

PhysicalActivitiesNo = +1.74 BMI (p < 2e-16)

Inactive people weigh almost 2 BMI units more, on average, than active people.
This is one of the strongest modifiable effects in the model.

<br>
**4. Sleep:** *Small negative effect*

SleepHours = –0.053 BMI units per hour (p = 0.0056)

Each additional hour of sleep is associated with a very small decrease in BMI.
Statistically significant but clinically minor.

<br>
**5. GeneralHealth:** *Strong monotonic gradient*

Compared with Excellent health, BMI increases as general health worsens:

- Very good: +1.64
- Good: +3.61
- Fair: +4.59
- Poor: +4.15
- All p < 2e-16.

This is a very clear and expected pattern:

Worse self-rated health → higher BMI.

These are among the largest effects in the model.

<br>
**6. SmokerStatus:** *Smoking groups have lower BMI than never-smokers*

Relative to the healthy baseline Never smoked:

- Current smoker (every day): –1.99 BMI (p < 2e-16)
- Current smoker (some days): –1.67 BMI (p < 2e-16)
- Former smoker: –0.018 (ns, p = 0.78)

This reproduces a well-established pattern:

Nicotine suppresses appetite, so current smokers tend to have lower BMI.

Former smokers resemble never-smokers, as weight is typically regained after quitting.

The strong negative coefficients confirm the reference level is now “Never smoked.”

<br>
**7. Model fit**

R² = 0.111 → 11.1% of BMI variance explained
This is typical for real-world lifestyle/health data, where BMI is influenced by many unmeasured factors (diet, genetics, socioeconomic conditions).

Residual SD ≈ 6.14 indicates substantial individual variability around predicted BMI.

F-test p < 2e-16 confirms the model as a whole is statistically meaningful.
</details>
<br>

<details>
<summary><b>Show plott: Linear Model BMI</b></summary>
```{r Plot lm BMI, fig.height=14, fig.width=8, cache=TRUE}
# Helper function
predict_by_factor <- function(lm_model, variable, baseline_list) {
  
  lvls <- levels(train[[variable]])
  
  newdata <- as.data.frame(baseline_list)
  newdata <- newdata[rep(1, length(lvls)), ]
  newdata[[variable]] <- lvls
  
  pred <- predict(lm_model, newdata = newdata, se.fit = TRUE)
  
  data.frame(
    Level = lvls,
    Fit   = pred$fit,
    Lower = pred$fit - 1.96 * pred$se.fit,
    Upper = pred$fit + 1.96 * pred$se.fit
  )
}

# Healthy baseline for BMI model
baseline_bmi <- list(
  AgeCategory        = "Age 18 to 24",
  Sex                = "Female",
  PhysicalActivities = "Yes",
  SleepHours         = mean(train$SleepHours, na.rm = TRUE),
  GeneralHealth      = "Excellent",
  SmokerStatus       = "Never smoked"
)

# Generate prediction tables
df_age  <- predict_by_factor(lm_bmi_healthy, "AgeCategory", baseline_bmi)
df_pa   <- predict_by_factor(lm_bmi_healthy, "PhysicalActivities", baseline_bmi)
df_gh   <- predict_by_factor(lm_bmi_healthy, "GeneralHealth", baseline_bmi)
df_smok <- predict_by_factor(lm_bmi_healthy, "SmokerStatus", baseline_bmi)

# Common y-axis across all panels
ymin_bmi <- min(df_age$Lower, df_pa$Lower, df_gh$Lower, df_smok$Lower)
ymax_bmi <- max(df_age$Upper, df_pa$Upper, df_gh$Upper, df_smok$Upper)

# Plot function (ordinal vs nominal)
plot_bmi_panel <- function(df, title, type = c("ordinal", "nominal")) {
  
  type <- match.arg(type)
  n <- nrow(df)
  xs <- seq(1, n)
  
  plot(xs, df$Fit, type="n",
       ylim = c(ymin_bmi, ymax_bmi),
       xlab = "", ylab = "Predicted BMI",
       main = title,
       xaxt = "n")
  
  # Grid
  grid(col="lightgray", lty=3, lwd=0.6)
  
  # Line ONLY for ordinal
  if (type == "ordinal") {
    lines(xs, df$Fit, lwd=2, col="blue")
  }
  
  # Points for all
  points(xs, df$Fit, pch=19, col="blue")
  
  # CI
  arrows(xs, df$Lower, xs, df$Upper, angle=90, code=3, length=0.05)
  
  axis(1, at = xs, labels = df$Level, las=2, cex.axis=0.8)
}

# Final 4 Plots
par(mfrow = c(4,1), mar = c(6,6,3,2))

plot_bmi_panel(df_age,  "Predicted BMI by Age Category",       "ordinal")
plot_bmi_panel(df_pa,   "Predicted BMI by Physical Activity",  "nominal")
plot_bmi_panel(df_gh,   "Predicted BMI by General Health",     "ordinal")
plot_bmi_panel(df_smok, "Predicted BMI by Smoking Status",     "nominal")

par(mfrow = c(1,1))
```
</details>
<br>

#### Linear Model 2: Predicting Sleep Hours

We do not need to log-transform SleepHours because the variable is already approximately continuous, bounded, and reasonably symmetric, and the residual diagnostics show no strong skewness or heteroscedasticity that would require transformation.
<br>

<details>
<summary><b>Show code: Linear Model Sleep Hours</b></summary>
```{r lm SleepHours, fig.height=14, fig.width=8, cache=TRUE}
### ======================================================
### 1. Relevel categorical predictors (logical order)
### ======================================================

# AgeCategory chronological
train$AgeCategory <- factor(train$AgeCategory, levels = c(
  "Age 18 to 24",
  "Age 25 to 29",
  "Age 30 to 34",
  "Age 35 to 39",
  "Age 40 to 44",
  "Age 45 to 49",
  "Age 50 to 54",
  "Age 55 to 59",
  "Age 60 to 64",
  "Age 65 to 69",
  "Age 70 to 74",
  "Age 75 to 79",
  "Age 80 or older"
))

# Physical activity logical
train$PhysicalActivities <- factor(train$PhysicalActivities,
                                   levels = c("Yes", "No"))

# General health ordered
train$GeneralHealth <- factor(train$GeneralHealth,
                              levels = c("Excellent",
                                         "Very good",
                                         "Good",
                                         "Fair",
                                         "Poor"))

# Smoking habits logical order
train$SmokerStatus <- factor(train$SmokerStatus,
                             levels = c("Never smoked",
                                        "Former smoker",
                                        "Current smoker - now smokes some days",
                                        "Current smoker - now smokes every day"))

# Sex baseline
train$Sex <- factor(train$Sex, levels = c("Female", "Male"))


### ======================================================
### 2. Fit LM model predicting SleepHours
### ======================================================

lm_sleep_healthy <- lm(
  SleepHours ~ AgeCategory + Sex + PhysicalActivities +
               GeneralHealth + DifficultyWalking + SmokerStatus,
  data = train
)

summary(lm_sleep_healthy)
```
</details>
<br>

Sleep duration varies systematically across age groups: adults aged 25–54 sleep about 0.1–0.25 hours less than those aged 18–24, whereas older adults sleep increasingly more, reaching +0.61 hours in those 80+. General health is strongly associated with sleep, with individuals in Poor health sleeping 0.49 hours less than those in Excellent health. Difficulty walking and current smoking show modest reductions of 0.12 hours and 0.10–0.14 hours, respectively, while sex and physical activity have no meaningful effects. Although the model’s explanatory power is limited, the observed associations provide clear patterns that align with the nonlinear effects later captured in the GAM.
<br>

<details>
<summary><b>Show detaild interpretation: Linear Model Sleep Hours</b></summary>
**This linear model estimates how demographic, lifestyle, mobility, and health characteristics are associated with nightly sleep duration. The reference individual — represented by the model intercept — is**

- Age 18–24
- Female
- Physically active (Yes)
- Excellent general health
- No difficulty walking
- Never smoked

Thus, the intercept of 7.15 hours represents the expected nightly sleep duration for a healthy young female, active, without mobility issues, and a non-smoker.

The coefficients represent additive deviations from this baseline.

**1. Age effects:** *Sleep decreases in early/mid-adulthood, then increases in older age*

Relative to the 18–24 baseline:

- Ages 25–54: sleep is 0.08–0.25 hours shorter, all statistically significant except 55–64.
- Ages 55–64: near zero difference (not significant).
- Ages 65–69: +0.24 hours (p < 1e-9)
- Ages 70–74: +0.35 hours
- Ages 75–79: +0.43 hours
- Age 80+: +0.61 hours, highly significant.

Pattern:
Sleep declines from young adulthood through midlife, then increases steadily beginning around retirement age.
This inverted U-shape is consistent with epidemiological evidence: midlife has the shortest sleep duration, older adults often sleep longer (though not always of higher quality).

**2. Sex:** *No meaningful difference*

SexMale = –0.011 hours, p = 0.43

This effect is tiny and not statistically significant.
→ There is no meaningful difference in sleep duration between males and females when adjusting for all other predictors.

**3. Physical Activity:** *Not significant*

PhysicalActivitiesNo = +0.028, p = 0.10

Inactive individuals sleep ~0.03 hours more, but the effect is very small and not statistically significant.
→ Physical activity does not meaningfully predict sleep duration in this dataset.

**4. General Health:** *Strong and monotonic association*

Compared with Excellent health, sleep duration decreases progressively with worsening health:

GeneralHealth level	Effect on SleepHours
Very good	–0.162 h
Good	–0.229 h
Fair	–0.351 h
Poor	–0.487 h

All effects are highly significant (p < 1e-13).

Interpretation:
Worse self-rated health is associated with materially shorter sleep.
Individuals reporting Poor health sleep nearly half an hour less per night than those in Excellent health.

This is one of the strongest predictors in the model.

**5. Difficulty Walking:** *Small but significant decrease*

DifficultyWalkingYes = –0.118 hours, p < 1e-7

People who report difficulty walking sleep about 7 minutes less per night on average.
This is statistically significant but clinically modest.

**6. Smoking:** *Sleep decreases among current smokers*

Relative to never-smokers:

- Former smokers: +0.026 h (ns)
- Current smokers (some days): –0.096 h, p = 0.01
- Current smokers (daily): –0.140 h, p < 1e-7

Interpretation:
Current smokers sleep 6–8 minutes less per night.
Former smokers sleep similarly to never-smokers.

This aligns with literature linking nicotine use to reduced sleep duration and poorer sleep quality.

**7. Model fit:**

Residual SD = 1.48 hours
→ Substantial unexplained variation, expected for behavioral outcomes like sleep.

R² = 0.037 (3.7%)
→ The model explains only a small portion of the variance.

F-test p < 2e-16
→ The model as a whole is statistically significant.

Interpretation:
Sleep duration is influenced by many unmeasured factors (stress, work hours, children, socioeconomic conditions, sleep disorders).
Even though R² is low, the direction and significance of effects provide meaningful insights into how sleep correlates with lifestyle and health indicators.
</details>
<br>

<br>
<details>
<summary><b>Show plot: Linear Model Sleep Hours</b></summary>
```{r Plot lm SleepHours, fig.height=14, fig.width=8, cache=TRUE}
### --- Helper to predict SleepHours for each level of a factor ---
predict_by_factor <- function(lm_model, variable, baseline_list) {
  
  lvls <- levels(train[[variable]])
  
  newdata <- as.data.frame(baseline_list)
  newdata <- newdata[rep(1, length(lvls)), ]
  newdata[[variable]] <- lvls
  
  pred <- predict(lm_model, newdata = newdata, se.fit = TRUE)
  
  data.frame(
    Level = lvls,
    Fit   = pred$fit,
    Lower = pred$fit - 1.96 * pred$se.fit,
    Upper = pred$fit + 1.96 * pred$se.fit
  )
}

# Healthy baseline for SleepHours model
baseline_sleep <- list(
  AgeCategory        = "Age 18 to 24",
  Sex                = "Female",
  PhysicalActivities = "Yes",
  GeneralHealth      = "Excellent",
  DifficultyWalking  = "No",
  SmokerStatus       = "Never smoked"
)

# Generate prediction tables
df_age_s   <- predict_by_factor(lm_sleep_healthy, "AgeCategory", baseline_sleep)
df_pa_s    <- predict_by_factor(lm_sleep_healthy, "PhysicalActivities", baseline_sleep)
df_gh_s    <- predict_by_factor(lm_sleep_healthy, "GeneralHealth", baseline_sleep)
df_smok_s  <- predict_by_factor(lm_sleep_healthy, "SmokerStatus", baseline_sleep)

# Common y-axis across all panels
y_min_s <- min(df_age_s$Lower, df_pa_s$Lower, df_gh_s$Lower, df_smok_s$Lower)
y_max_s <- max(df_age_s$Upper, df_pa_s$Upper, df_gh_s$Upper, df_smok_s$Upper)

# Plotting function (ordinal ↔ nominal logic built in)
plot_sleep_panel <- function(df, title, variable_type = c("ordinal","nominal")) {
  
  variable_type <- match.arg(variable_type)
  n <- nrow(df)
  xs <- seq(1, n)
  
  plot(xs, df$Fit, type="n",
       ylim = c(y_min_s, y_max_s),
       xlab = "", ylab = "Predicted Sleep Hours",
       main = title,
       xaxt = "n")
  
  grid(col="lightgray", lty=3, lwd=0.6)
  
  # Ordinal → draw a line
  if (variable_type == "ordinal") {
    lines(xs, df$Fit, lwd=2, col="blue")
  }
  
  # Points for all
  points(xs, df$Fit, pch=19, col="blue")
  arrows(xs, df$Lower, xs, df$Upper, angle=90, code=3, length=0.05)
  
  axis(1, at = xs, labels = df$Level, las=2, cex.axis=0.9)
}

# Final 4 stacked plots
par(mfrow = c(4,1), mar = c(6,6,3,2))

plot_sleep_panel(df_age_s,  "Predicted Sleep Hours by Age Category",      "ordinal")
plot_sleep_panel(df_pa_s,   "Predicted Sleep Hours by Physical Activity", "nominal")
plot_sleep_panel(df_gh_s,   "Predicted Sleep Hours by General Health",    "ordinal")
plot_sleep_panel(df_smok_s, "Predicted Sleep Hours by Smoking Status",    "nominal")

par(mfrow = c(1,1))
```
</details>
<br>

### Generalized Linear Model Poisson

Julia

### Generalized Additive Model

Samuel
```{r gam-binomial, cache=TRUE}
library(mgcv)

gam_bin <- bam(
  HadHeartAttack ~
    s(BMI, k = 10) +
    s(SleepHours, k = 10) +
    AgeCategory + Sex + SmokerStatus +
    HadDiabetes + PhysicalActivities + GeneralHealth +
    DifficultyWalking + HadKidneyDisease + CovidPos,
  data = train,
  family = binomial,
  method = "fREML"
)

summary(gam_bin)
```

```{r gam-plots, fig.width=7, fig.height=6, cache=TRUE}
plot(gam_bin, pages = 1, shade = TRUE, seWithMean = TRUE, scale = 0)
```


<br>

--------------------------------------------------------------------------------

<br>

# 4. Results and Discussion

Focus on main results
• Tables and figures should provide evidence for main result
• Precise and correct desription and interpretation of main results

## 3.1. Risk prediction: Can a model reliably identify patients at increased risk?

GLM Binomial (Samuel)

SVM (Luca)

NN (Julia)

## 3.2. Risk explanation: Which factors drive the risk, and how do these effects unfold?

LM (Luca)

GLM Poisson (Julia)

GAM (Samuel)

<br>

--------------------------------------------------------------------------------

<br>

# 5. Conclusion

• Summary of most important findings
• Put findings into a broader context
• Ideas for further research

<br>

--------------------------------------------------------------------------------
