---
title: "Early detection and risk stratification of heart disease"
author: "Luca Signorini, Samuel Frei, Julia Studer"
date: "09.01.2025"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
---

# Introduction

**Background**

This project was completed as part of the Master’s Program in Applied Information and Data Science at HSLU in the module Applied Machine Learning and Predictive Modelling 1 and forms part of the final course assessment.

**The objective** 

The objective of this project is to **analyze and model the occurrence of self-reported heart attacks using a structured health survey dataset** comprising demographic characteristics, lifestyle factors, and medical history variables. Rather than focusing on causal inference, the analysis aims to evaluate the predictive performance of different modeling approaches and to assess how selected risk factors contribute to model predictions.

**Methodology**

Given the strong class imbalance in the response variable HadHeartAttack, appropriate preprocessing steps are applied to construct balanced training data. An exploratory data analysis is conducted to obtain an initial understanding of the distributions of predictors such as body mass index (BMI) and sleep duration, as well as their relationship with the response variable.

Multiple classification models are subsequently trained and compared, including a logistic regression model as a baseline, as well as more flexible approaches such as generalized additive models, support vector machines and neural networks. Model performance is primarily evaluated using receiver operating characteristic (ROC) curves and the area under the curve (AUC).

The results are interpreted with respect to predictive accuracy and the direction and magnitude of estimated effects, with particular attention to the limitations imposed by the observational nature of the data and the use of self-reported health information.

**Motivation & Use-case**

Cardiovascular diseases represent a major public health concern and are associated with a wide range of demographic, behavioral, and health-related factors. Large-scale health survey datasets provide an opportunity to apply statistical and machine learning methods to examine how such factors relate to the occurrence of cardiovascular events and to evaluate their usefulness for risk prediction.

As a motivating use case, the estimated risk scores produced by the models are considered in the context of prioritization rather than diagnosis. In a hospital or screening-like setting, such scores could be used to rank individuals by predicted risk, enabling threshold-dependent strategies that trade off sensitivity and specificity according to operational objectives.

This use case is intended purely as contextual motivation. The analysis is based on observational, self-reported survey data and does not support causal inference or clinical decision-making. Accordingly, all results are interpreted strictly in terms of predictive performance and relative risk ranking, not as medical diagnosis or treatment guidance.

--------------------------------------------------------------------------------

# Data Preparation

**Dataset:** heart_2022_no_nans.csv
<br>
**Source:** https://www.kaggle.com/datasets/kamilpytlak/personal-key-indicators-of-heart-disease?resource=download&select=2022 

**Background Dataset**

The chosen data set is an excerpt of the data collected in 2022 in the framework of the Behavioral Risk Factor Surveillance System (BRFSS) of the american Centers for Disease Control and Prevention (CDC). In this framework, people are called randomly with Random Digit Dialing techniques on landlines and cellphones and are asked to answer a questionnaire regarding their health-related risk behaviors and events, chronic health conditions and use of preventive services. The survey is conducted contituously through the year and adults 18 years and older are asked to participate. For each question there are some options to choose from like Yes/No; number of days 1-30; good, bad, fair. In some cases people are asked for numbers without suggestions, e.g. weight and height for the Body-Mass-Index (BMI) calculation. 
After loading the data set, we reduce both the number of variables and observations to stay within the task limits. 

(Sources: https://www.cdc.gov/brfss/about/brfss_faq.htm, https://www.cdc.gov/brfss/questionnaires/pdf-ques/2022-BRFSS-Questionnaire-508.pdf). 

**Handling Class Imbalance in the Target Variable**

In the original dataset, the outcome variable HadHeartAttack is highly imbalanced, with only a small fraction of participants reporting a past heart attack. This imbalance poses a major challenge for predictive models: Without correction, most models learn to simply predict “No heart attack” for all cases. While this can still produce a high overall accuracy, such models fail exactly where the hospital needs them most — identifying patients at increased cardiovascular risk.

Due to the large sample size of the provided dataset, we were required to use the undersampling method to achieve a 4:1 ratio of our target variable - 'HadHeartAttack' - as well as to meet the dataset size prerequisites. From an initial sample size of over 200'000, we reduced it to about 67'000, of which 53'740 did not have a heart attack and 13'435 did. After this we split the remaining Set in a Training and Testing Set of ratio 70:30.

<br>
<details>
<summary><b>Show code: Prerequisites</b></summary>
```{r prerequisites}
# Prerequisites
#install.packages("dplyr")
library(dplyr)
#install.packages("skimr")
library(skimr)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("nnet")
library(nnet)
#install.packages("gamlss.add")
library(gamlss.add)
#install.packages("ggplot2")
library(ggplot2)
#install.library("caret")
library(caret)
#install.packages("PRROC")
library(PRROC)
#install.packages("pROC")
library(pROC)
#install.library("ROCR")
library(ROCR)
#install.library("caret")
library(caret)
#install.packages("ggrepel")
library(ggrepel)
```
</details>

<details>
<summary><b>Show code: Data import and summary</b></summary>
```{r loading-data, cache=TRUE}
# Loading data
#data <- read.csv("heart_2022_no_nans.csv")
data <- read.csv(file = "/Users/lucasignorini/Documents/MSc_ids/Sem_3/ML1/Gruppenarbeit/Data/archive/2022/heart_2022_no_nans.csv")

skim(data)
```

```{r recode-outcome, cache=TRUE}
# Ensure outcome is a proper binary factor
data$HadHeartAttack <- factor(data$HadHeartAttack,
                              levels = c("No", "Yes"))

# Quick check
str(data$HadHeartAttack)
table(data$HadHeartAttack)
```
</details>

<details>
<summary><b>Show code: Convert-types</b></summary>
```{r convert-types, cache=TRUE}
data <- data %>%
  mutate(
    HadHeartAttack     = factor(HadHeartAttack, levels = c("No", "Yes")),
    AgeCategory        = factor(AgeCategory),
    Sex                = factor(Sex),
    SmokerStatus       = factor(SmokerStatus),
    HadDiabetes        = factor(HadDiabetes),
    PhysicalActivities = factor(PhysicalActivities),
    GeneralHealth      = factor(GeneralHealth),
    DifficultyWalking  = factor(DifficultyWalking),
    HadKidneyDisease   = factor(HadKidneyDisease),
    CovidPos           = factor(CovidPos),
    BMI                = as.numeric(BMI),
    SleepHours         = as.numeric(SleepHours)
  )

str(data$AgeCategory)
```
</details>

<details>
<summary><b>Show code: Select predictors</b></summary>
```{r select-predictors, cache=TRUE}
# Keep only 11 selected predictors = 12 variables
df <- data %>%
  select(
    HadHeartAttack,
    AgeCategory,
    Sex,
    BMI,
    SmokerStatus,
    HadDiabetes,
    PhysicalActivities,
    GeneralHealth,
    SleepHours,
    DifficultyWalking,
    HadKidneyDisease,
    CovidPos
  )

```
</details>

<details>
<summary><b>Show code: Handling class imbalance</b></summary>
```{r handling-class-imbalance, cache=TRUE}

set.seed(123)

minority <- df %>% filter(HadHeartAttack == "Yes")
majority <- df %>% filter(HadHeartAttack == "No")

ratio <- 4  # majority : minority = 4 : 1

majority_under <- majority %>% sample_n(nrow(minority) * ratio)

df_balanced <- bind_rows(minority, majority_under)

# Overwrite name so colleague's code still works:
data <- df_balanced

# Check new distribution
table(data$HadHeartAttack)
```
</details>

<details>
<summary><b>Show code: Split Data in Test and Train Set</b></summary>
```{r split, cache=TRUE}
set.seed(123)

train_index <- createDataPartition(data$HadHeartAttack, p = 0.7, list = FALSE)

train <- data[train_index, ]
test  <- data[-train_index, ]


```
</details>
--------------------------------------------------------------------------------

<br>

# EDA

In order to get a first impression of the training dataset, we make some general plots.

**Response imbalance**

```{r,EDA Response imbalance, cache=TRUE}
barplot(table(train$HadHeartAttack), main="HadHeartAttack class counts", ylab="Count")
```

**Predictor distribution**

```{r,EDA Predictor distribution, cache=TRUE}
hist(train$BMI, breaks = 40, main = "BMI distribution", xlab = "BMI")
abline(v = mean(train$BMI, na.rm = TRUE), col = "red", lwd = 2, lty = 2)
m <- mean(train$BMI, na.rm = TRUE)
text(x = m,
     y = 6500,
     labels = paste("Mean =", round(m, 2)),
     pos = 4,
     col = "red")

hist(train$SleepHours, breaks = seq(0.5, 24.5, by=1), main="SleepHours distribution", xlab="Sleep hours")
```

**BMI:**
The BMI distribution is right-skewed, with most observations concentrated between roughly 20 and 35, and a long tail toward higher BMI values.
The mean BMI of 28.79 lies in the overweight range, indicating that the average individual in the dataset has a BMI above the healthy reference interval.

**Sleeping hours:**
The distribution of sleep hours is approximately centered around 7–8 hours, with most individuals reporting sleep durations within this range.
The distribution shows slight right skewness, with relatively few observations at very low or very high sleep durations.

**BMI by heart attack**
```{r,EDA BMI by heart attack, cache=TRUE}
boxplot(BMI ~ HadHeartAttack, data=train,
        main="BMI by HadHeartAttack", xlab="HadHeartAttack", ylab="BMI")
```

**BMI by heartattak:**
Individuals who reported a prior heart attack show a slightly higher median BMI compared to those without a heart attack.
However, the substantial overlap between the two distributions indicates that BMI alone provides limited discriminatory power between the groups.

--------------------------------------------------------------------------------

# Risk prediction: Can a model reliably identify patients at increased risk?

This section focuses on the predictive task of identifying individuals at increased risk of having experienced a heart attack based on demographic, lifestyle, and health-related variables. The aim is not causal inference, but to assess how reliably different modeling approaches can discriminate between individuals with and without a reported heart attack.

A binomial generalized linear model (GLM) is first applied as an interpretable baseline, providing a reference for predictive performance under standard statistical assumptions. Subsequently, more flexible machine learning models, including a support vector machine (SVM) and a neural network (NN), are trained to evaluate whether increased model complexity leads to improved predictive accuracy. Model performance is compared using appropriate classification metrics, with particular emphasis on ROC curves and AUC.

## Generalized linear model Binomial
<details>
<summary><b>Show Code & Output: Generalized linear model Binomial </b></summary>
```{r glm-binomial, cache=TRUE}
glm_bin <- glm(
  HadHeartAttack ~ AgeCategory + Sex + BMI + SmokerStatus +
    HadDiabetes + PhysicalActivities + GeneralHealth +
    SleepHours + DifficultyWalking + HadKidneyDisease + CovidPos,
  data = train,
  family = binomial
)

summary(glm_bin)
```
</details>
<br>

Age is the strongest driver in the baseline GLM: compared to the reference (youngest) group, coefficients increase monotonically across age categories, indicating substantially higher odds of a reported heart attack with increasing age.

Self-rated general health is highly informative: relative to the reference level, worse health categories (e.g., Fair / Poor) show large positive coefficients, reflecting markedly higher predicted odds and strong discriminative value.

Sex shows a clear association: SexMale is strongly positive (β ≈ 0.80; OR ≈ 2.23), meaning males have higher predicted odds than females when holding other predictors constant.

Disclaimer: These are associations from observational, self-reported data and should not be interpreted causally. We do not focus on single-variable interpretation; instead, we highlight a few representative effects to illustrate that the baseline model behaves in a realistic and expected way before comparing predictive performance across models.

<details>
<summary><b>Show Code: GLM Binomial ROC </b></summary>
```{r glm-roc, cache=TRUE}
library(pROC)

prob_glm <- predict(glm_bin, newdata = test, type = "response")

roc_glm <- roc(test$HadHeartAttack, prob_glm)

plot(roc_glm, main = "ROC Curve — GLM Binomial")
auc(roc_glm)
```
</details>
<br>

The ROC curve shows clear separation from the diagonal, indicating good discriminatory ability of the baseline GLM.
An AUC of 0.8249 suggests that the model reliably distinguishes between individuals with and without a reported heart attack and provides a strong baseline for further comparison.

<details>
<summary><b>Show Code: GLM Binomial BMI-curve </b></summary>
```{r glm-bmi-curve, cache=TRUE}
bmi_grid <- data.frame(
  BMI                = seq(15, 50, length.out = 150),
  AgeCategory        = levels(train$AgeCategory)[1],
  Sex                = levels(train$Sex)[1],
  SmokerStatus       = levels(train$SmokerStatus)[1],
  HadDiabetes        = levels(train$HadDiabetes)[1],
  PhysicalActivities = levels(train$PhysicalActivities)[1],
  GeneralHealth      = levels(train$GeneralHealth)[1],
  SleepHours         = 7,
  DifficultyWalking  = levels(train$DifficultyWalking)[1],
  HadKidneyDisease   = levels(train$HadKidneyDisease)[1],
  CovidPos           = levels(train$CovidPos)[1]
)

bmi_grid$prob <- predict(glm_bin, newdata = bmi_grid, type = "response")

plot(
  bmi_grid$BMI, bmi_grid$prob, type = "l", lwd = 2,
  xlab = "BMI", ylab = "Predicted probability",
  main = "Predicted Heart Attack Risk vs BMI (GLM)"
)
```
</details>
<br>

The predicted risk increases only marginally with BMI in the binomial GLM, and the corresponding coefficient is not statistically significant, indicating a weak linear association.
Although exploratory analysis suggests small differences across BMI ranges, this example illustrates that a strictly linear specification may be insufficient and motivates the use of more flexible, non-linear models to capture potential effects.

--------------------------------------------------------------------------------

## Support Vector Machines

Support Vector Machines offer a powerful alternative to traditional statistical models by allowing nonlinear decision boundaries through kernel functions. In the context of heart disease risk prediction, nonlinearities are plausible: risk factors such as age, BMI, diabetes, or general health may interact in complex ways that are not fully captured by a general linear model as above. An SVM with a radial basis function (RBF) kernel can automatically model such interactions without needing to specify them explicitly.

Our goal in applying an SVM is therefore twofold:

- Evaluate whether nonlinear patterns improve prediction accuracy relative to the GLM baseline.
- Assess the model’s ability to assign individual risk probabilities, which can support clinical risk stratification.

Because RBF SVMs scale poorly to large datasets (quadratic to cubic computational complexity), we train the model on a 10,000-row subsample and use cross-validated hyperparameter tuning to optimize its performance. This approach provides a computationally feasible way to test whether more flexible nonlinear models meaningfully improve risk prediction over interpretable linear models.
<br>

<details>
<summary><b>Show code: SVM Model </b></summary>
```{r, SVM Model, cache=TRUE}
### ---------------------------------------------------------------
### Support Vector Machine (SVM) for Heart Attack Risk Prediction
### Radial kernel + hyperparameter tuning (caret / kernlab)
### Evaluation: ROC/AUC + PR curve/AUPRC + Confusion Matrix
### ---------------------------------------------------------------

# Ensure factor levels (control = "No", case = "Yes")
train$HadHeartAttack <- factor(train$HadHeartAttack, levels = c("No", "Yes"))
test$HadHeartAttack  <- factor(test$HadHeartAttack,  levels = c("No", "Yes"))

# -----------------------------------------------------------
# 1) Stratified subsample for computational feasibility
# -----------------------------------------------------------
set.seed(123)
sub_index <- createDataPartition(
  y = train$HadHeartAttack,
  p = 10000 / nrow(train),
  list = FALSE
)
train_sub <- train[sub_index, ]

# -----------------------------------------------------------
# 2) Hyperparameter tuning via grid search + 3-fold CV
# -----------------------------------------------------------
tune_grid <- expand.grid(
  sigma = c(0.005, 0.01, 0.02, 0.05),
  C     = c(0.25, 0.5, 1, 2)
)

ctrl <- trainControl(
  method = "cv",
  number = 3,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  allowParallel = TRUE
)

set.seed(123)
svm_tuned <- train(
  HadHeartAttack ~ AgeCategory + Sex + BMI + SmokerStatus +
    HadDiabetes + PhysicalActivities + GeneralHealth +
    SleepHours + DifficultyWalking + HadKidneyDisease + CovidPos,
  data = train_sub,
  method = "svmRadial",
  metric = "ROC",
  tuneGrid = tune_grid,
  trControl = ctrl,
  preProcess = c("center", "scale")
)
```
</details>

<details>
<summary><b>Show Output: SVM Model, tuned sigma and C</b></summary>
```{r, SVM_tuned, cache=TRUE}
svm_tuned  # shows best sigma/C by ROC
```
</details>

```{r, SVM model evaluation, cache=TRUE, echo=FALSE}
# -----------------------------------------------------------
# 3) Predict probabilities on the test set
# -----------------------------------------------------------
svm_prob <- predict(svm_tuned, newdata = test, type = "prob")[, "Yes"]

# -----------------------------------------------------------
# 4) ROC curve + AUC
# -----------------------------------------------------------
roc_svm <- roc(
  response  = test$HadHeartAttack,
  predictor = svm_prob,
  levels    = c("No", "Yes"),
  direction = "<"
)

plot(
  roc_svm,
  main = "ROC Curve — SVM (Radial, tuned)",
  legacy.axes = TRUE,   # x-axis becomes 1 - specificity
  xlim = c(1, 0),
  ylim = c(0, 1)
)

auc(roc_svm)

# -----------------------------------------------------------
# 5) Precision–Recall (PR) curve + AUPRC
# -----------------------------------------------------------
scores_pos <- svm_prob[test$HadHeartAttack == "Yes"]
scores_neg <- svm_prob[test$HadHeartAttack == "No"]

pr_svm <- pr.curve(
  scores.class0 = scores_pos,
  scores.class1 = scores_neg,
  curve = TRUE
)

plot(pr_svm, main = "PR Curve — SVM (Radial, tuned)")
paste("Area under PR curve: ", round(pr_svm$auc.integral, digits = 4)) 

```

<details>
<summary><b>Show code: SVM model evaluation </b></summary>
```{html, SVM model evaluation Dropdown, cache=TRUE}
# -----------------------------------------------------------
# 3) Predict probabilities on the test set
# -----------------------------------------------------------
svm_prob <- predict(svm_tuned, newdata = test, type = "prob")[, "Yes"]

# -----------------------------------------------------------
# 4) ROC curve + AUC
# -----------------------------------------------------------
roc_svm <- roc(
  response  = test$HadHeartAttack,
  predictor = svm_prob,
  levels    = c("No", "Yes"),
  direction = "<"
)

plot(
  roc_svm,
  main = "ROC Curve — SVM (Radial, tuned)",
  legacy.axes = TRUE,   # x-axis becomes 1 - specificity
  xlim = c(1, 0),
  ylim = c(0, 1)
)

auc(roc_svm)

# -----------------------------------------------------------
# 5) Precision–Recall (PR) curve + AUPRC
# -----------------------------------------------------------
scores_pos <- svm_prob[test$HadHeartAttack == "Yes"]
scores_neg <- svm_prob[test$HadHeartAttack == "No"]

pr_svm <- pr.curve(
  scores.class0 = scores_pos,
  scores.class1 = scores_neg,
  curve = TRUE
)

plot(pr_svm, main = "PR Curve — SVM (Radial, tuned)")
pr_svm$auc.integral   # AUPRC (area under PR curve)

```
</details>
<br>

**Model interpretation**

The SVM’s predictive performance is assessed using ROC and Precision–Recall (PR) curves.
The ROC curve shows that the model has good discriminative ability (AUC ≈ 0.80), meaning it can generally rank patients with a heart attack higher than those without.

The PR curve, which is more informative under class imbalance, reveals a rapid decline in precision as recall increases, resulting in a moderate AUPRC (≈ 0.48). This indicates that while the model can identify high-risk individuals, increased sensitivity is accompanied by a substantial number of false positives.

**Parameter Interpretation**

The radial SVM is controlled by the kernel width σ and the regularization parameter C.
Lower values of σ produce smoother decision boundaries, while lower C values penalize misclassifications less strongly and favor conservative models.

The selected parameters (σ = 0.005, C = 0.25) indicate a smooth, regularized model optimized for overall discrimination (ROC), rather than aggressive identification of positive cases. As a result, the model emphasizes stable probability estimates, making threshold selection a key determinant of classification performance.

<br>
```{r, SVM Risk distribution histogram and Confusion Matrix for two thresholds, cache=TRUE, echo=FALSE}
# -----------------------------------------------------------
# Risk Distribution for SVM showing the two treshholds
# -----------------------------------------------------------
plot_df <- data.frame(Risk = svm_prob)

ggplot(plot_df, aes(x = Risk)) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white",
    alpha = 0.7
  ) +

  # Threshold line at 0.2
  geom_vline(
    xintercept = 0.2,
    linetype = 1,
    linewidth = 1,
    color = "red"
  ) +

  # Label for the threshold
  annotate(
    "text",
    x = 0.35,
    y = 9250,
    label = "Decision threshold = 0.20",
    size = 3.5,
    color = "red"
  ) +
  
    # Threshold line at 0.5
  geom_vline(
    xintercept = 0.5,
    linetype = 2,
    linewidth = 1,
    color = "gray"
  ) +

  # Label for the threshold
  annotate(
    "text",
    x = 0.65,
    y = 9250,
    label = "Decision threshold = 0.50",
    size = 3.5,
    color = "gray"
  ) +

  labs(
    title = "Risk Distribution — SVM-Predicted Heart Attack Probability",
    x = "Predicted probability of heart attack",
    y = "Number of individuals"
  ) +
  theme_minimal(base_size = 13)

# -----------------------------------------------------------
# Confusion Matrix at threshold 0.5 (default operating point)
# -----------------------------------------------------------
svm_pred_class_moderate <- factor(ifelse(svm_prob >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
svm_pred_class_lenient <- factor(ifelse(svm_prob >= 0.2, "Yes", "No"), levels = c("No", "Yes"))

confusionMatrix(
  data      = svm_pred_class_moderate,
  reference = test$HadHeartAttack,
  positive  = "Yes"
)

confusionMatrix(
  data      = svm_pred_class_lenient,
  reference = test$HadHeartAttack,
  positive  = "Yes"
)

```

<br>
<details>
<summary><b>Show code: SVM Risk distribution histogram and Confusion Matrix for two thresholds </b></summary>
```{html, SVM Risk distribution histogram and Confusion Matrix for two thresholds Dropdown}
# -----------------------------------------------------------
# Risk Distribution for SVM showing the two treshholds
# -----------------------------------------------------------
plot_df <- data.frame(Risk = svm_prob)

ggplot(plot_df, aes(x = Risk)) +
  geom_histogram(
    bins = 40,
    fill = "steelblue",
    color = "white",
    alpha = 0.7
  ) +

  # Threshold line at 0.2
  geom_vline(
    xintercept = 0.2,
    linetype = 1,
    linewidth = 1,
    color = "red"
  ) +

  # Label for the threshold
  annotate(
    "text",
    x = 0.35,
    y = 9250,
    label = "Decision threshold = 0.20",
    size = 3.5,
    color = "red"
  ) +
  
    # Threshold line at 0.5
  geom_vline(
    xintercept = 0.5,
    linetype = 2,
    linewidth = 1,
    color = "gray"
  ) +

  # Label for the threshold
  annotate(
    "text",
    x = 0.65,
    y = 9250,
    label = "Decision threshold = 0.50",
    size = 3.5,
    color = "gray"
  ) +

  labs(
    title = "Risk Distribution — SVM-Predicted Heart Attack Probability",
    x = "Predicted probability of heart attack",
    y = "Number of individuals"
  ) +
  theme_minimal(base_size = 13)

# -----------------------------------------------------------
# Confusion Matrix at threshold 0.5 (default operating point)
# -----------------------------------------------------------
svm_pred_class_moderate <- factor(ifelse(svm_prob >= 0.5, "Yes", "No"), levels = c("No", "Yes"))
svm_pred_class_lenient <- factor(ifelse(svm_prob >= 0.2, "Yes", "No"), levels = c("No", "Yes"))

confusionMatrix(
  data      = svm_pred_class_moderate,
  reference = test$HadHeartAttack,
  positive  = "Yes"
)

confusionMatrix(
  data      = svm_pred_class_lenient,
  reference = test$HadHeartAttack,
  positive  = "Yes"
)

```
</details>
<br>

**Risk distribution & Threshold-dependent classification performance**

The risk distribution (histogram) supplements the ROC and PR curves with an intuitive view of the risk values that the SVM model actually outputs in the test data set. It can be seen that the majority of individuals receive a low predicted probability (strong peak in the left-hand area), while only a small proportion have a significantly increased modeled probability (right-hand edge/long “tail”).

Because the model outputs probabilities, it does not provide one single “best” classification by itself — we must choose a threshold that reflects the clinical goal (e.g., avoid missed cases vs. avoid false alarms). We evaluated two operating points:

- *Moderate threshold:* t = 0.5 (stricter; fewer “Yes” predictions)

With t=0.5, the SVM behaves like a high-confidence alert system: it produces relatively few positive predictions (Detection Prevalence ~5.8%), and most non-cases are correctly filtered out. However, it misses the majority of true cases (FN is very large), which is problematic for early detection.

- *Lenient threshold:* t = 0.2 (more sensitive; more “Yes” predictions)

Lowering the threshold to t=0.2 shifts the model towards a screening / early-warning behavior: it identifies substantially more at-risk patients (higher sensitivity), at the cost of more false positives. This is typically the preferred direction when the main goal is not missing high-risk individuals, provided the healthcare system can handle additional follow-up checks.

The prevalence in the test set is about 20% (“Yes”: 0.19998). This matters because accuracy can look high even if a model mostly predicts “No”.

**Conclusion**

The tuned radial SVM provides meaningful risk ranking and outputs individual probabilities that can support clinical risk stratification. However, in this dataset it does not improve overall discrimination compared with the GLM baseline: the SVM reaches an ROC–AUC of about 0.80, while the GLM achieves 0.82. Thus, the expected benefit from nonlinear modeling is limited here.

The model’s practical usefulness depends primarily on threshold choice. With a lower threshold (e.g., t = 0.2), the SVM behaves as an early-warning tool—identifying more at-risk patients at the cost of additional false positives—appropriate when follow-up assessment is feasible. A higher threshold (e.g., t = 0.5) yields fewer alerts but misses many true cases.

The selected hyperparameters (σ = 0.005, C = 0.25) favor a smoother decision boundary optimized for ROC performance; alternative choices (e.g., tuning with PR-focused criteria) could shift the balance toward higher recall for the positive class. Overall, the SVM is suitable for probability-based risk stratification, but requires careful thresholding aligned with clinical goals.
<br>

--------------------------------------------------------------------------------

## Neural Network 

For the first model the same parameters as in the German Credit example from the lecture was used on the training data defined earlier.  

<details>
<summary>Show output</summary>
```{r, Neural Network model, cache=TRUE}
#the train proportion of the data defined above is used to train the model

set.seed(412)
HeartAttack_net <- nnet(HadHeartAttack ~ ., data = train, size=(16), maxit=10000, range=0.1, decay=5e-4, MaxNWts = 2000)
```
</details>

```{r,echo=TRUE, results='hide'}
HeartAttack_net
```
The model converged after 690 iterations.  

We are mostly interested in correctly predicting the heart attack cases correctly. It could be fatal if a patient is labeled no risk for heart attack that actually has a high risk. On the other hand it is not so tragic if a person with moderate risk is labelled as high risk, this could motivate a healthier lifestyle. 

### Make predictions

```{r, cache=TRUE} 
pred <- predict(HeartAttack_net, test, type="class")

confusionMatrix(as.factor(pred), as.factor(test$HadHeartAttack))
```
The model is very good at identifying the "No heart attack" cases with sensitivity value of 0.944. At the same time the model misses 68% of heart attack cases, wit a specificity value of 0.321. This is not enough for a risk prediction in medical modeling. This is also represented in the ROC curve shown below.     

### ROC Curves

The Receiver Operating Characteristic is a useful tool to indicate model quality. For this we plot sensitivity against 1-specificity.

```{r,ROC curves, cache=TRUE}

pred_raw <- predict(HeartAttack_net, test, decision.values=TRUE, type = "raw")
pred <- ROCR::prediction(pred_raw, test$HadHeartAttack)
perf <- ROCR::performance(pred, "tpr", "fpr")
plot(perf, lwd=2, col="blue")
abline(a=0, b=1)
```


The main reasons for this are the following: first we have a high class imbalance (80/20), even tough undersampling was applied to the original data. Also the default classification threshold is 0.5, which is not appropriate in this case. The model as it is optimised for accuracy, which gives more weight to the correct prediction of the most cases of no heart attack. 

### Optimised neural network model

To correct for these factors, the following parameters were changed:

1. Scale predictors 
The scaling of the predictors should improve the prediction of the HadHeartAttack Yes case. Also it keeps the neural network weights stable and the threshold tuning reliable.`

2. Rebalance the training data 
Since there are still mostly "HadHeartAttack No" cases used for training, the "HadHeartAttack Yes" cases are upsampled.

3. Optimizing the parameters of the model
There are imbalanced classes and missing a heart attack is worse than a false alarm and therefore accuracy is misleading. The ROC AUC is optimized and probabilities used. The main parameters to be optimized are the size and decay. 

<details>
<summary>Show analysis</summary>
```{r,nn analysis}

# Define cross-validation strategy
#ctrl <- trainControl(
  #method = "repeatedcv",
  #number = 5,
  #repeats = 3,
  #sampling = "up",          # handle imbalance inside CV
  #classProbs = TRUE,        # needed for ROC
  #summaryFunction = twoClassSummary,
  #savePredictions = "final"
#)

# Define the tuning grid 
#grid <- expand.grid(
  #size = c(4, 8, 12),
  #decay = c(1e-4, 5e-4, 1e-3, 5e-3)
#)

# Train and tune the model

#set.seed(412)

#nnet_caret <- train(
  #HadHeartAttack ~ .,
  #data = train,
  #method = "nnet",
  #metric = "ROC",          # optimize AUC
  #tuneGrid = grid,
  #trControl = ctrl,
  #preProcess = c("center", "scale"),
  #trace = FALSE,
  #maxit = 1000
#)

#plot(HeartAttack_v2_net)
```
</details>

The model HeartAttack_v2_net was then trained with the optimized parameters.
<br>

<details>
<summary>Show output</summary>
```{r, cache=TRUE}
# Inspect optimal parameters
#nnet_caret

#nnet_caret$bestTune
```
```{r}
# Train the model with the optimized parameters
set.seed(412)
HeartAttack_v2_net <- nnet(HadHeartAttack ~ ., data = train, size=(4), maxit=10000, range=0.1, decay=1e-4, MaxNWts = 2000)
```
</details>

Now the model converges after 320 iterations. It then is used to make predictions on the test part of the dataset. 

### Make predictions

```{r, cache=TRUE} 
pred_v2 <- predict(HeartAttack_v2_net, test, type="raw")

pred_raw <- predict(HeartAttack_net, test, decision.values=TRUE, type = "raw")
pred <- ROCR::prediction(pred_raw, test$HadHeartAttack)
perf <- ROCR::performance(pred, "tpr", "fpr")
plot(perf, lwd=2, col="blue")
abline(a=0, b=1)
```

```{r, Area under curve ROC optimized, cache=TRUE}  

library(pROC)
pred_class <- ifelse(pred_v2 > 0.2, "Yes", "No")
pred_class <- factor(pred_class, levels = c("No", "Yes"))

confusionMatrix(
  pred_class,
  test$HadHeartAttack,
  positive = "Yes"
)

roc_obj <- roc(test$HadHeartAttack, pred_v2)
coords(roc_obj, x = c(0.05, 0.1, 0.15, 0.20, 0.25, 0.30),
       input = "threshold",
       ret = c("sensitivity", "specificity"))
``` 

Looking at the results of using different thresholds, there is a trade off between high sensitivity (correct detection of HadHeartAttack Yes) and specificity (correct dismissing of HadHeartAttack No). A good compromise of detecting the risk of a heart attack and not have too many false alarms, is a threshold of 0.2 or 0.15. Looking at the balanced accuracy, the maximum is at a threshold of 0.2, balancing best the two error types. Like this, the correct predicted HadHeartAttack are 79 %, and the correct dismiss of HadHeartAttack Yes is still at 71 %. Also a threshold of 0.15 could be applied if the false positives are considered as acceptable. 

<br>

----------------------------------------------------------

<br>

# Regression-Based Modeling

## Exploratory Linear Models Among Predictors

To better understand how the health indicators relate to one another — independently of the heart attack outcome — we fitted simple linear models using BMI and SleepHours as continuous response variables. These exploratory models help reveal clusters of risk factors and provide context for the effects observed in the GLM and GAM models.
<br>

### Predicting BMI

We did not log-transform the BMI do it is a continous variable because it is strictly positive, just slightly right-skewed and most importantly the interpretation of the Model-Output remains intuitiv.

<br>
<details>
<summary><b>Show code: linear model predicting BMI</b></summary>
```{r lm_bim, cache=TRUE}
### ======================================================
### ORDER FACTORS IN A LOGICAL WAY
### ======================================================

# Age (chronological)
train$AgeCategory <- factor(train$AgeCategory, levels = c(
  "Age 18 to 24",
  "Age 25 to 29",
  "Age 30 to 34",
  "Age 35 to 39",
  "Age 40 to 44",
  "Age 45 to 49",
  "Age 50 to 54",
  "Age 55 to 59",
  "Age 60 to 64",
  "Age 65 to 69",
  "Age 70 to 74",
  "Age 75 to 79",
  "Age 80 or older"
))

# Ordered health categories
train$GeneralHealth <- factor(train$GeneralHealth,
                              levels = c("Excellent", "Very good", "Good", "Fair", "Poor"))

# Physical activity (nominal)
train$PhysicalActivities <- factor(train$PhysicalActivities,
                                   levels = c("Yes", "No"))

# Smoking (nominal)
train$SmokerStatus <- factor(train$SmokerStatus,
                             levels = c("Never smoked",
                                        "Former smoker",
                                        "Current smoker - now smokes some days",
                                        "Current smoker - now smokes every day"))

# Sex
train$Sex <- factor(train$Sex, levels = c("Female", "Male"))


### ======================================================
### FIT THE MODEL WITH HEALTHY BASELINE
### ======================================================

lm_bmi_healthy <- lm(
  BMI ~ AgeCategory + Sex + PhysicalActivities + SleepHours +
        GeneralHealth + SmokerStatus,
  data = train
)

```
</details>
<br>

<details>
<summary>Show output summary</summary>
```{r lm_bim output, cache=TRUE}
summary(lm_bmi_healthy)
```
</details>
<br>

Using a healthy reference profile (young, female, active, excellent health, never smoked), the model reveals clear associations with BMI. 

BMI rises steadily across adulthood and peaks around ages 45–55, consistent with epidemiological patterns, before declining in the oldest age groups. Physical inactivity is associated with a substantial BMI increase of 1.7 units. Worse self-rated general health shows a strong monotonic association with higher BMI, with individuals reporting “Fair” or “Poor” health exhibiting more than 4 BMI units above the healthy baseline. Current smokers have significantly lower BMI than never-smokers, a known effect of nicotine. Although the model explains around 11% of BMI variation—typical for behavioral health data—it provides meaningful insight into how lifestyle and health characteristics cluster in ways relevant for cardiovascular risk.

<details>
<summary><b>Show detailed interpretation: linear model predicting BMI</b></summary>
<br>
**This linear model estimates how demographic, lifestyle, and health variables are associated with BMI, using a healthy individual as the reference category:**

- Age 18–24
- Female
- Physically active (PhysicalActivities = Yes)
- Excellent general health
- Never smoked

The intercept (23.84) is therefore the estimated BMI for this baseline individual.

<br>
**1. Age effects:** *BMI increases steadily with age until ~55–59, then declines in very old age*

Relative to the 18–24 baseline:

- Ages 25–29: +2.00 BMI
- Ages 40–44: +3.77 BMI
- Ages 45–49: +4.05 BMI
- Ages 50–54: +3.98 BMI
- Ages 55–59: +3.39 BMI

These are all highly significant (p < 2e-16).

This pattern reflects a typical adult BMI trajectory:
BMI rises through early and middle adulthood, stabilizes, then decreases at the oldest ages.

Age 80+ shows –0.47 BMI (p = 0.0049) versus young adults → older adults often experience muscle loss and weight reduction.

<br>
**2. Sex:** *Males have slightly higher BMI*

SexMale = +0.27 BMI units (p ≈ 2.6e-06)

This indicates a modest but significant difference, consistent with population norms.

<br>
**3. Physical activity:** *Inactive individuals have substantially higher BMI*

PhysicalActivitiesNo = +1.74 BMI (p < 2e-16)

Inactive people weigh almost 2 BMI units more, on average, than active people.
This is one of the strongest modifiable effects in the model.

<br>
**4. Sleep:** *Small negative effect*

SleepHours = –0.053 BMI units per hour (p = 0.0056)

Each additional hour of sleep is associated with a very small decrease in BMI.
Statistically significant but clinically minor.

<br>
**5. GeneralHealth:** *Strong monotonic gradient*

Compared with Excellent health, BMI increases as general health worsens:

- Very good: +1.64
- Good: +3.61
- Fair: +4.59
- Poor: +4.15
- All p < 2e-16.

This is a very clear and expected pattern:

Worse self-rated health → higher BMI.

These are among the largest effects in the model.

<br>
**6. SmokerStatus:** *Smoking groups have lower BMI than never-smokers*

Relative to the healthy baseline Never smoked:

- Current smoker (every day): –1.99 BMI (p < 2e-16)
- Current smoker (some days): –1.67 BMI (p < 2e-16)
- Former smoker: –0.018 (ns, p = 0.78)

This reproduces a well-established pattern:

Nicotine suppresses appetite, so current smokers tend to have lower BMI.

Former smokers resemble never-smokers, as weight is typically regained after quitting.

The strong negative coefficients confirm the reference level is now “Never smoked.”

<br>
**7. Model fit**

R² = 0.111 → 11.1% of BMI variance explained
This is typical for real-world lifestyle/health data, where BMI is influenced by many unmeasured factors (diet, genetics, socioeconomic conditions).

Residual SD ≈ 6.14 indicates substantial individual variability around predicted BMI.

F-test p < 2e-16 confirms the model as a whole is statistically meaningful.
</details>
<br>

```{r, Plot lm BMI, fig.height=14, fig.width=8, cache=TRUE, echo=FALSE}
# Helper function
predict_by_factor <- function(lm_model, variable, baseline_list) {
  
  lvls <- levels(train[[variable]])
  
  newdata <- as.data.frame(baseline_list)
  newdata <- newdata[rep(1, length(lvls)), ]
  newdata[[variable]] <- lvls
  
  pred <- predict(lm_model, newdata = newdata, se.fit = TRUE)
  
  data.frame(
    Level = lvls,
    Fit   = pred$fit,
    Lower = pred$fit - 1.96 * pred$se.fit,
    Upper = pred$fit + 1.96 * pred$se.fit
  )
}

# Healthy baseline for BMI model
baseline_bmi <- list(
  AgeCategory        = "Age 18 to 24",
  Sex                = "Female",
  PhysicalActivities = "Yes",
  SleepHours         = mean(train$SleepHours, na.rm = TRUE),
  GeneralHealth      = "Excellent",
  SmokerStatus       = "Never smoked"
)

# Generate prediction tables
df_age  <- predict_by_factor(lm_bmi_healthy, "AgeCategory", baseline_bmi)
df_pa   <- predict_by_factor(lm_bmi_healthy, "PhysicalActivities", baseline_bmi)
df_gh   <- predict_by_factor(lm_bmi_healthy, "GeneralHealth", baseline_bmi)
df_smok <- predict_by_factor(lm_bmi_healthy, "SmokerStatus", baseline_bmi)

# Common y-axis across all panels
ymin_bmi <- min(df_age$Lower, df_pa$Lower, df_gh$Lower, df_smok$Lower)
ymax_bmi <- max(df_age$Upper, df_pa$Upper, df_gh$Upper, df_smok$Upper)

# Plot function (ordinal vs nominal)
plot_bmi_panel <- function(df, title, type = c("ordinal", "nominal")) {
  
  type <- match.arg(type)
  n <- nrow(df)
  xs <- seq(1, n)
  
  plot(xs, df$Fit, type="n",
       ylim = c(ymin_bmi, ymax_bmi),
       xlab = "", ylab = "Predicted BMI",
       main = title,
       xaxt = "n")
  
  # Grid
  grid(col="lightgray", lty=3, lwd=0.6)
  
  # Line ONLY for ordinal
  if (type == "ordinal") {
    lines(xs, df$Fit, lwd=2, col="blue")
  }
  
  # Points for all
  points(xs, df$Fit, pch=19, col="blue")
  
  # CI
  arrows(xs, df$Lower, xs, df$Upper, angle=90, code=3, length=0.05)
  
  axis(1, at = xs, labels = df$Level, las=2, cex.axis=0.8)
}

# Final 4 Plots
par(mfrow = c(4,1), mar = c(6,6,3,2))

plot_bmi_panel(df_age,  "Predicted BMI by Age Category",       "ordinal")
plot_bmi_panel(df_pa,   "Predicted BMI by Physical Activity",  "nominal")
plot_bmi_panel(df_gh,   "Predicted BMI by General Health",     "ordinal")
plot_bmi_panel(df_smok, "Predicted BMI by Smoking Status",     "nominal")

par(mfrow = c(1,1))
```


<details>
<summary><b>Show Code: Linear Model BMI Plotts</b></summary>
```{html, Plot lm BMI Dropdown, fig.height=14, fig.width=8, cache=TRUE}
# Helper function
predict_by_factor <- function(lm_model, variable, baseline_list) {
  
  lvls <- levels(train[[variable]])
  
  newdata <- as.data.frame(baseline_list)
  newdata <- newdata[rep(1, length(lvls)), ]
  newdata[[variable]] <- lvls
  
  pred <- predict(lm_model, newdata = newdata, se.fit = TRUE)
  
  data.frame(
    Level = lvls,
    Fit   = pred$fit,
    Lower = pred$fit - 1.96 * pred$se.fit,
    Upper = pred$fit + 1.96 * pred$se.fit
  )
}

# Healthy baseline for BMI model
baseline_bmi <- list(
  AgeCategory        = "Age 18 to 24",
  Sex                = "Female",
  PhysicalActivities = "Yes",
  SleepHours         = mean(train$SleepHours, na.rm = TRUE),
  GeneralHealth      = "Excellent",
  SmokerStatus       = "Never smoked"
)

# Generate prediction tables
df_age  <- predict_by_factor(lm_bmi_healthy, "AgeCategory", baseline_bmi)
df_pa   <- predict_by_factor(lm_bmi_healthy, "PhysicalActivities", baseline_bmi)
df_gh   <- predict_by_factor(lm_bmi_healthy, "GeneralHealth", baseline_bmi)
df_smok <- predict_by_factor(lm_bmi_healthy, "SmokerStatus", baseline_bmi)

# Common y-axis across all panels
ymin_bmi <- min(df_age$Lower, df_pa$Lower, df_gh$Lower, df_smok$Lower)
ymax_bmi <- max(df_age$Upper, df_pa$Upper, df_gh$Upper, df_smok$Upper)

# Plot function (ordinal vs nominal)
plot_bmi_panel <- function(df, title, type = c("ordinal", "nominal")) {
  
  type <- match.arg(type)
  n <- nrow(df)
  xs <- seq(1, n)
  
  plot(xs, df$Fit, type="n",
       ylim = c(ymin_bmi, ymax_bmi),
       xlab = "", ylab = "Predicted BMI",
       main = title,
       xaxt = "n")
  
  # Grid
  grid(col="lightgray", lty=3, lwd=0.6)
  
  # Line ONLY for ordinal
  if (type == "ordinal") {
    lines(xs, df$Fit, lwd=2, col="blue")
  }
  
  # Points for all
  points(xs, df$Fit, pch=19, col="blue")
  
  # CI
  arrows(xs, df$Lower, xs, df$Upper, angle=90, code=3, length=0.05)
  
  axis(1, at = xs, labels = df$Level, las=2, cex.axis=0.8)
}

# Final 4 Plots
par(mfrow = c(4,1), mar = c(6,6,3,2))

plot_bmi_panel(df_age,  "Predicted BMI by Age Category",       "ordinal")
plot_bmi_panel(df_pa,   "Predicted BMI by Physical Activity",  "nominal")
plot_bmi_panel(df_gh,   "Predicted BMI by General Health",     "ordinal")
plot_bmi_panel(df_smok, "Predicted BMI by Smoking Status",     "nominal")

par(mfrow = c(1,1))
```
</details>
<br>

--------------------------------------------------------------------------------

### Predicting Sleep Hours

We do not need to log-transform SleepHours because the variable is already approximately continuous, bounded, and reasonably symmetric, and the residual diagnostics show no strong skewness or heteroscedasticity that would require transformation.
<br>

<details>
<summary><b>Show code: Linear Model Sleep Hours</b></summary>
```{r lm SleepHours, fig.height=14, fig.width=8, cache=TRUE}
### ======================================================
### 1. Relevel categorical predictors (logical order)
### ======================================================

# AgeCategory chronological
train$AgeCategory <- factor(train$AgeCategory, levels = c(
  "Age 18 to 24",
  "Age 25 to 29",
  "Age 30 to 34",
  "Age 35 to 39",
  "Age 40 to 44",
  "Age 45 to 49",
  "Age 50 to 54",
  "Age 55 to 59",
  "Age 60 to 64",
  "Age 65 to 69",
  "Age 70 to 74",
  "Age 75 to 79",
  "Age 80 or older"
))

# Physical activity logical
train$PhysicalActivities <- factor(train$PhysicalActivities,
                                   levels = c("Yes", "No"))

# General health ordered
train$GeneralHealth <- factor(train$GeneralHealth,
                              levels = c("Excellent",
                                         "Very good",
                                         "Good",
                                         "Fair",
                                         "Poor"))

# Smoking habits logical order
train$SmokerStatus <- factor(train$SmokerStatus,
                             levels = c("Never smoked",
                                        "Former smoker",
                                        "Current smoker - now smokes some days",
                                        "Current smoker - now smokes every day"))

# Sex baseline
train$Sex <- factor(train$Sex, levels = c("Female", "Male"))


### ======================================================
### 2. Fit LM model predicting SleepHours
### ======================================================

lm_sleep_healthy <- lm(
  SleepHours ~ AgeCategory + Sex + PhysicalActivities +
               GeneralHealth + DifficultyWalking + SmokerStatus,
  data = train
)

```
</details>
<br>

<details>
<summary>Show output summary</summary>
```{r lm SleepHours Output, fig.height=14, fig.width=8, cache=TRUE}
summary(lm_sleep_healthy)
```
</details>
<br>

Sleep duration varies systematically across age groups: adults aged 25–54 sleep about 0.1–0.25 hours less than those aged 18–24, whereas older adults sleep increasingly more, reaching +0.61 hours in those 80+. General health is strongly associated with sleep, with individuals in Poor health sleeping 0.49 hours less than those in Excellent health. Difficulty walking and current smoking show modest reductions of 0.12 hours and 0.10–0.14 hours, respectively, while sex and physical activity have no meaningful effects. Although the model’s explanatory power is limited, the observed associations provide clear patterns that align with the nonlinear effects later captured in the GAM.
<br>

<details>
<summary><b>Show detaild interpretation: Linear Model Sleep Hours</b></summary>
**This linear model estimates how demographic, lifestyle, mobility, and health characteristics are associated with nightly sleep duration. The reference individual — represented by the model intercept — is**

- Age 18–24
- Female
- Physically active (Yes)
- Excellent general health
- No difficulty walking
- Never smoked

Thus, the intercept of 7.15 hours represents the expected nightly sleep duration for a healthy young female, active, without mobility issues, and a non-smoker.

The coefficients represent additive deviations from this baseline.

**1. Age effects:** *Sleep decreases in early/mid-adulthood, then increases in older age*

Relative to the 18–24 baseline:

- Ages 25–54: sleep is 0.08–0.25 hours shorter, all statistically significant except 55–64.
- Ages 55–64: near zero difference (not significant).
- Ages 65–69: +0.24 hours (p < 1e-9)
- Ages 70–74: +0.35 hours
- Ages 75–79: +0.43 hours
- Age 80+: +0.61 hours, highly significant.

Pattern:
Sleep declines from young adulthood through midlife, then increases steadily beginning around retirement age.
This inverted U-shape is consistent with epidemiological evidence: midlife has the shortest sleep duration, older adults often sleep longer (though not always of higher quality).

**2. Sex:** *No meaningful difference*

SexMale = –0.011 hours, p = 0.43

This effect is tiny and not statistically significant.
→ There is no meaningful difference in sleep duration between males and females when adjusting for all other predictors.

**3. Physical Activity:** *Not significant*

PhysicalActivitiesNo = +0.028, p = 0.10

Inactive individuals sleep ~0.03 hours more, but the effect is very small and not statistically significant.
→ Physical activity does not meaningfully predict sleep duration in this dataset.

**4. General Health:** *Strong and monotonic association*

Compared with Excellent health, sleep duration decreases progressively with worsening health:

GeneralHealth level	Effect on SleepHours
Very good	–0.162 h
Good	–0.229 h
Fair	–0.351 h
Poor	–0.487 h

All effects are highly significant (p < 1e-13).

Interpretation:
Worse self-rated health is associated with materially shorter sleep.
Individuals reporting Poor health sleep nearly half an hour less per night than those in Excellent health.

This is one of the strongest predictors in the model.

**5. Difficulty Walking:** *Small but significant decrease*

DifficultyWalkingYes = –0.118 hours, p < 1e-7

People who report difficulty walking sleep about 7 minutes less per night on average.
This is statistically significant but clinically modest.

**6. Smoking:** *Sleep decreases among current smokers*

Relative to never-smokers:

- Former smokers: +0.026 h (ns)
- Current smokers (some days): –0.096 h, p = 0.01
- Current smokers (daily): –0.140 h, p < 1e-7

Interpretation:
Current smokers sleep 6–8 minutes less per night.
Former smokers sleep similarly to never-smokers.

This aligns with literature linking nicotine use to reduced sleep duration and poorer sleep quality.

**7. Model fit:**

Residual SD = 1.48 hours
→ Substantial unexplained variation, expected for behavioral outcomes like sleep.

R² = 0.037 (3.7%)
→ The model explains only a small portion of the variance.

F-test p < 2e-16
→ The model as a whole is statistically significant.

Interpretation:
Sleep duration is influenced by many unmeasured factors (stress, work hours, children, socioeconomic conditions, sleep disorders).
Even though R² is low, the direction and significance of effects provide meaningful insights into how sleep correlates with lifestyle and health indicators.
</details>
<br>

```{r Plot lm SleepHours, fig.height=14, fig.width=8, cache=TRUE, echo=FALSE}
### --- Helper to predict SleepHours for each level of a factor ---
predict_by_factor <- function(lm_model, variable, baseline_list) {
  
  lvls <- levels(train[[variable]])
  
  newdata <- as.data.frame(baseline_list)
  newdata <- newdata[rep(1, length(lvls)), ]
  newdata[[variable]] <- lvls
  
  pred <- predict(lm_model, newdata = newdata, se.fit = TRUE)
  
  data.frame(
    Level = lvls,
    Fit   = pred$fit,
    Lower = pred$fit - 1.96 * pred$se.fit,
    Upper = pred$fit + 1.96 * pred$se.fit
  )
}

# Healthy baseline for SleepHours model
baseline_sleep <- list(
  AgeCategory        = "Age 18 to 24",
  Sex                = "Female",
  PhysicalActivities = "Yes",
  GeneralHealth      = "Excellent",
  DifficultyWalking  = "No",
  SmokerStatus       = "Never smoked"
)

# Generate prediction tables
df_age_s   <- predict_by_factor(lm_sleep_healthy, "AgeCategory", baseline_sleep)
df_pa_s    <- predict_by_factor(lm_sleep_healthy, "PhysicalActivities", baseline_sleep)
df_gh_s    <- predict_by_factor(lm_sleep_healthy, "GeneralHealth", baseline_sleep)
df_smok_s  <- predict_by_factor(lm_sleep_healthy, "SmokerStatus", baseline_sleep)

# Common y-axis across all panels
y_min_s <- min(df_age_s$Lower, df_pa_s$Lower, df_gh_s$Lower, df_smok_s$Lower)
y_max_s <- max(df_age_s$Upper, df_pa_s$Upper, df_gh_s$Upper, df_smok_s$Upper)

# Plotting function (ordinal ↔ nominal logic built in)
plot_sleep_panel <- function(df, title, variable_type = c("ordinal","nominal")) {
  
  variable_type <- match.arg(variable_type)
  n <- nrow(df)
  xs <- seq(1, n)
  
  plot(xs, df$Fit, type="n",
       ylim = c(y_min_s, y_max_s),
       xlab = "", ylab = "Predicted Sleep Hours",
       main = title,
       xaxt = "n")
  
  grid(col="lightgray", lty=3, lwd=0.6)
  
  # Ordinal → draw a line
  if (variable_type == "ordinal") {
    lines(xs, df$Fit, lwd=2, col="blue")
  }
  
  # Points for all
  points(xs, df$Fit, pch=19, col="blue")
  arrows(xs, df$Lower, xs, df$Upper, angle=90, code=3, length=0.05)
  
  axis(1, at = xs, labels = df$Level, las=2, cex.axis=0.9)
}

# Final 4 stacked plots
par(mfrow = c(4,1), mar = c(6,6,3,2))

plot_sleep_panel(df_age_s,  "Predicted Sleep Hours by Age Category",      "ordinal")
plot_sleep_panel(df_pa_s,   "Predicted Sleep Hours by Physical Activity", "nominal")
plot_sleep_panel(df_gh_s,   "Predicted Sleep Hours by General Health",    "ordinal")
plot_sleep_panel(df_smok_s, "Predicted Sleep Hours by Smoking Status",    "nominal")

par(mfrow = c(1,1))
```


<details>
<summary><b>Show Code: Linear Model Sleep Hours Plots</b></summary>
```{html, Plot lm SleepHours Dropdown, fig.height=14, fig.width=8, cache=TRUE}
### --- Helper to predict SleepHours for each level of a factor ---
predict_by_factor <- function(lm_model, variable, baseline_list) {
  
  lvls <- levels(train[[variable]])
  
  newdata <- as.data.frame(baseline_list)
  newdata <- newdata[rep(1, length(lvls)), ]
  newdata[[variable]] <- lvls
  
  pred <- predict(lm_model, newdata = newdata, se.fit = TRUE)
  
  data.frame(
    Level = lvls,
    Fit   = pred$fit,
    Lower = pred$fit - 1.96 * pred$se.fit,
    Upper = pred$fit + 1.96 * pred$se.fit
  )
}

# Healthy baseline for SleepHours model
baseline_sleep <- list(
  AgeCategory        = "Age 18 to 24",
  Sex                = "Female",
  PhysicalActivities = "Yes",
  GeneralHealth      = "Excellent",
  DifficultyWalking  = "No",
  SmokerStatus       = "Never smoked"
)

# Generate prediction tables
df_age_s   <- predict_by_factor(lm_sleep_healthy, "AgeCategory", baseline_sleep)
df_pa_s    <- predict_by_factor(lm_sleep_healthy, "PhysicalActivities", baseline_sleep)
df_gh_s    <- predict_by_factor(lm_sleep_healthy, "GeneralHealth", baseline_sleep)
df_smok_s  <- predict_by_factor(lm_sleep_healthy, "SmokerStatus", baseline_sleep)

# Common y-axis across all panels
y_min_s <- min(df_age_s$Lower, df_pa_s$Lower, df_gh_s$Lower, df_smok_s$Lower)
y_max_s <- max(df_age_s$Upper, df_pa_s$Upper, df_gh_s$Upper, df_smok_s$Upper)

# Plotting function (ordinal ↔ nominal logic built in)
plot_sleep_panel <- function(df, title, variable_type = c("ordinal","nominal")) {
  
  variable_type <- match.arg(variable_type)
  n <- nrow(df)
  xs <- seq(1, n)
  
  plot(xs, df$Fit, type="n",
       ylim = c(y_min_s, y_max_s),
       xlab = "", ylab = "Predicted Sleep Hours",
       main = title,
       xaxt = "n")
  
  grid(col="lightgray", lty=3, lwd=0.6)
  
  # Ordinal → draw a line
  if (variable_type == "ordinal") {
    lines(xs, df$Fit, lwd=2, col="blue")
  }
  
  # Points for all
  points(xs, df$Fit, pch=19, col="blue")
  arrows(xs, df$Lower, xs, df$Upper, angle=90, code=3, length=0.05)
  
  axis(1, at = xs, labels = df$Level, las=2, cex.axis=0.9)
}

# Final 4 stacked plots
par(mfrow = c(4,1), mar = c(6,6,3,2))

plot_sleep_panel(df_age_s,  "Predicted Sleep Hours by Age Category",      "ordinal")
plot_sleep_panel(df_pa_s,   "Predicted Sleep Hours by Physical Activity", "nominal")
plot_sleep_panel(df_gh_s,   "Predicted Sleep Hours by General Health",    "ordinal")
plot_sleep_panel(df_smok_s, "Predicted Sleep Hours by Smoking Status",    "nominal")

par(mfrow = c(1,1))
```
</details>
<br>

--------------------------------------------------------------------------------

## Predicting sleeping hours with GLM Poisson


Here, the factor of the sleeping hours is investigated more in detail. For this, the effect of the age category, sex, physical activity smoker status and general health on the sleeping hours is examined.
The variable SleepHours is rather a measured duration of time reported in whole hours and not a cout of events, so it is not a count variable in a strict definition. Nevertheless it is reported as a count-like outcome with non-negative integers and is here treated as a count variable.

<details>
<summary><b>Show code: Relevel predictors</b></summary>
``` {r gam-poisson, cache=TRUE}

### ======================================================
### 1. Relevel categorical predictors (logical order)
### ======================================================

# AgeCategory chronological
train$AgeCategory <- factor(train$AgeCategory, levels = c(
  "Age 18 to 24",
  "Age 25 to 29",
  "Age 30 to 34",
  "Age 35 to 39",
  "Age 40 to 44",
  "Age 45 to 49",
  "Age 50 to 54",
  "Age 55 to 59",
  "Age 60 to 64",
  "Age 65 to 69",
  "Age 70 to 74",
  "Age 75 to 79",
  "Age 80 or older"
))

# Physical activity logical
train$PhysicalActivities <- factor(train$PhysicalActivities,
                                   levels = c("Yes", "No"))

# General health ordered
train$GeneralHealth <- factor(train$GeneralHealth,
                              levels = c("Excellent",
                                         "Very good",
                                         "Good",
                                         "Fair",
                                         "Poor"))

# Smoking habits logical order
train$SmokerStatus <- factor(train$SmokerStatus,
                             levels = c("Never smoked",
                                        "Former smoker",
                                        "Current smoker - now smokes some days",
                                        "Current smoker - now smokes every day"))

# Sex baseline
train$Sex <- factor(train$Sex, levels = c("Female", "Male"))


str(data)
data.sleep <- data |> 
  mutate(
    AgeCategory = factor(AgeCategory, ordered = TRUE),
    SleepHours = as.integer(SleepHours)
  )
str(data.sleep)

boxplot(
  SleepHours ~ AgeCategory, 
  data = data.sleep,
  las = 2, 
  main = "Sleep Hours by Age Category",
  ylab = "Sleep hours"
)

boxplot(
  SleepHours ~ factor(PhysicalActivities, ordered = TRUE), 
  data = data.sleep,
  las = 2, 
  main = "Sleep Hours by Physical activities",
  ylab = "Sleep hours"
)

boxplot(
  SleepHours ~ GeneralHealth, 
  data = data.sleep,
  las = 2, 
  main = "Sleep Hours by General Health",
  ylab = "Sleep hours"
)
boxplot(
  SleepHours ~ SmokerStatus, 
  data = data.sleep,
  las = 2, 
  main = "Sleep Hours by Smoker Status",
  ylab = "Sleep hours"
)
boxplot(
  SleepHours ~ Sex, 
  data = data.sleep,
  las = 2, 
  main = "Sleep Hours by Sex",
  ylab = "Sleep hours"
)
```
</details>
<br>

<details>
<summary><b>Show code: GLM poisson model</b></summary>
```{r}

#3. Fit a Poisson Model 
#Model Specification

glm.sleep <- glm(
  SleepHours ~
    AgeCategory +
    Sex +
    BMI +
    SmokerStatus +
    HadDiabetes +
    PhysicalActivities +
    GeneralHealth +
    DifficultyWalking +
    HadKidneyDisease +
    CovidPos,
  data = df,
  family = "poisson"
)
coef(glm.sleep)

summary(glm.sleep)
```
</details>
<br>

After adjusting for health and lifestyle factors, predicted mean sleep hours differ modestly across the population:

Age: Younger adults (25–39) sleep ~2–3% less than the youngest group, whereas older adults (65–80+) sleep 3–8% more.

General health: Those with poor health sleep ~6% fewer hours than those with excellent health (~0.4 hours difference).

Smoking and BMI: Former and never smokers sleep slightly more than daily smokers (~2–3% difference), and higher BMI is associated with a minimal decrease (~0.06% per BMI unit).

Sex: Males sleep only ~0.5% less than females.
Overall, differences are statistically significant but practically modest; age and general health have the largest effects.

If we look at the dispersion of 0.29 we find a strong underdispersion. This is mainly due to the sleep hours being bounded (1-24) and the responses highly heaped around 6,7,8 hours. Also the large sample produces very stable conditional means. So the Poisson variance assumption (Var = Mean) is too flexible for this constrained data. This results in the standard errors being conservative and the p-values larger than they should be. Assumptions from Proisson model are not violated in a dangerous way. The model is not corrected and Quasi-Poisson is not appropriate for underdispersion. 

<details>
<summary><b>Show code: Generate Age & Main Predictor Plots</b></summary>
```{r, age-and-main-plots, echo=TRUE, message=FALSE, warning=FALSE}
#----------------------------------------
# 1. Add predicted values from the Poisson model
#----------------------------------------
df$pred_sleep <- predict(glm.sleep, type = "response")

#----------------------------------------
# 2. Summary function for predictors with 95% CI
#----------------------------------------
summarise_by_predictor <- function(data, predictor, predictor_name) {
  data %>%
    group_by({{ predictor }}) %>%
    summarise(
      n         = n(),
      obs_mean  = mean(SleepHours, na.rm = TRUE),
      obs_se    = sd(SleepHours, na.rm = TRUE)/sqrt(n),
      obs_lwr   = obs_mean - 1.96*obs_se,
      obs_upr   = obs_mean + 1.96*obs_se,
      pred_mean = mean(pred_sleep, na.rm = TRUE),
      .groups   = "drop"
    ) %>%
    rename(level = {{ predictor }}) %>%
    mutate(predictor = predictor_name)
}

#----------------------------------------
# 3. Build summary datasets
#----------------------------------------
sum_age <- summarise_by_predictor(df, AgeCategory, "Age category")

sum_main <- bind_rows(
  summarise_by_predictor(df, GeneralHealth, "General health"),
  summarise_by_predictor(df, Sex, "Sex"),
  summarise_by_predictor(df, SmokerStatus, "Smoking status"),
  summarise_by_predictor(df, PhysicalActivities, "Physical activity")
)

predictors <- unique(sum_main$predictor)

#----------------------------------------
# 4. Set axis limits for plots
#----------------------------------------
axis_lim_age  <- range(c(sum_age$obs_mean, sum_age$pred_mean), na.rm = TRUE)
axis_lim_main <- range(c(sum_main$obs_mean, sum_main$pred_mean), na.rm = TRUE)

margin_age <- 0.5
axis_lim_age  <- c(axis_lim_age[1]-margin_age, axis_lim_age[2]+margin_age)
margin_main <- 0.5
axis_lim_main <- c(axis_lim_main[1]-margin_main, axis_lim_main[2]+margin_main)

#----------------------------------------
# 5. Plot: AgeCategory
#----------------------------------------
age_palette_13 <- c(
  "#000000", "#E69F00", "#56B4E9", "#009E73",
  "#F0E442", "#0072B2", "#D55E00", "#CC79A7",
  "#999999", "#0099CC", "#66CC33", "#CC0000", "#9933FF"
)

p_age <- ggplot(sum_age, aes(x = obs_mean, y = pred_mean)) +
  geom_errorbarh(aes(xmin = obs_lwr, xmax = obs_upr), height = 0.03, color = "grey60") +
  geom_point(aes(color = level), size = 3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  coord_equal(xlim = axis_lim_age, ylim = axis_lim_age, clip = "off") +
  labs(
    x = "Observed mean sleep hours (95% CI)",
    y = "Predicted mean sleep hours",
    title = "Observed vs Predicted Mean Sleep Hours by Age Category",
    color = "Age category"
  ) +
  scale_color_manual(values = age_palette_13) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5), plot.margin = margin(t=25, r=25, b=5, l=20))

#----------------------------------------
# 6. Plot: Main predictors (single panels)
#----------------------------------------
plots_list <- list()

for(pred in predictors){
  df_pred <- subset(sum_main, predictor == pred)

  p <- ggplot(df_pred, aes(x = obs_mean, y = pred_mean)) +
    geom_errorbarh(aes(xmin = obs_lwr, xmax = obs_upr), height = 0.03, color = "blue") +
    geom_point(color = "red", size = 3) +
    geom_text_repel(aes(label = level),
                    size = 3,
                    min.segment.length = 0,
                    box.padding = 0.6,
                    point.padding = 0.4,
                    force = 2,
                    max.overlaps = Inf,
                    seed = 123) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
    coord_equal(xlim = axis_lim_main, ylim = axis_lim_main, clip = "off") +
    labs(
      x = "Observed mean sleep hours (95% CI)",
      y = "Predicted mean sleep hours",
      title = paste("Observed vs Predicted Mean Sleep Hours:", pred)
    ) +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(hjust = 0.5), plot.margin = margin(t=20, r=20, b=20, l=20))

  plots_list[[pred]] <- p
}
```
</details>

```{r,show plots} 

p_age # Age plot

for(p in plots_list) print(p) # Main predictor plots
```

<br>

-----------------------------------------------------

<br>

## Generalized Additive Model 

<details>
<summary>Show Code & Summary: GAM </summary>
```{r gam-binomial, cache=TRUE}
library(mgcv)

gam_bin <- bam(
  HadHeartAttack ~
    s(BMI, k = 10) +
    s(SleepHours, k = 10) +
    AgeCategory + Sex + SmokerStatus +
    HadDiabetes + PhysicalActivities + GeneralHealth +
    DifficultyWalking + HadKidneyDisease + CovidPos,
  data = train,
  family = binomial,
  method = "fREML"
)

summary(gam_bin)
```
</details>
<br>

To assess whether nonlinear modeling improves risk prediction, we compared the binomial GLM with a binomial GAM using the same set of predictors. While the GAM allows flexible nonlinear effects for continuous variables such as BMI and sleep duration, it did not lead to a clear improvement in predictive performance compared to the GLM. The binomial GLM already achieved strong discrimination on the test set (AUC ≈ 0.83), outperforming the more complex machine learning models considered in this study, such as the tuned SVM.

This suggests that, given the available predictors, heart attack risk follows a largely additive and approximately linear structure on the log-odds scale. As a result, the simpler GLM provides an excellent balance between predictive accuracy, stability, and interpretability, making it the preferred model for clinical risk screening.

To directly compare predictive performance, we evaluated both the binomial GLM and the binomial GAM using ROC curves on the test set. The GLM achieved an AUC of approximately 0.83, while the GAM yielded a very similar AUC, indicating no meaningful improvement in discrimination when allowing nonlinear smooth effects for BMI and sleep duration.

From a risk prediction perspective, this result suggests that the additional flexibility of the GAM is not required to reliably identify high-risk individuals in this dataset. The simpler GLM therefore remains preferable for screening purposes due to its comparable performance, lower complexity, and clearer interpretability.

```{r gam-plots, fig.width=7, fig.height=6, cache=TRUE}
plot(gam_bin, pages = 1, shade = TRUE, seWithMean = TRUE, scale = 0)
```

Although the GAM does not substantially improve predictive performance over the GLM, it provides important explanatory advantages. By allowing nonlinear smooth effects particularly for sleep duration, and to a lesser extent for BMI, the GAM reveals how risk changes across the full range of these variables rather than assuming a constant linear effect.

In particular, sleep duration shows a clear nonlinear association with heart attack risk, indicating that risk is not minimized at extreme values but rather within an intermediate range. This type of pattern cannot be captured by a standard GLM without imposing ad-hoc transformations. The GAM therefore plays a complementary role: it helps identify risk thresholds and clinically relevant ranges, supporting targeted prevention and patient counseling.

<br>

--------------------------------------------------------------------------------

<br>

# Results and Discussion

**Model performance**

The binomial GLM achieved the highest and most stable predictive performance on the test set with a ROC–AUC of approximately 0.82. The tuned radial SVM reached a slightly lower discrimination (ROC–AUC ≈ 0.80) and therefore did not outperform the baseline. The binomial GAM achieved comparable ROC performance to the GLM, indicating no improvement in overall discrimination from introducing nonlinear smooth terms.

For models that output probabilities (SVM and NN), overall performance depends not only on discrimination but also on the chosen classification threshold. In the context of heart attack risk prediction, where missed cases are particularly costly, lower thresholds improve sensitivity. For the Neural Network, a threshold around 0.2 yields the best balance between sensitivity and specificity, achieving the highest balanced accuracy. If false positives are acceptable, an even lower threshold (e.g., 0.15) may be preferred to further reduce the risk of missed heart attack cases.

**Effect modeling**

While the GLM assumes linear effects on the log-odds scale, the GAM revealed nonlinear effect shapes, particularly for sleep duration and to a lesser extent for BMI. However, these nonlinearities did not translate into measurable gains in predictive accuracy.

**Threshold behavior**

Most evaluated models produce probability scores rather than fixed classifications. Performance therefore varies substantially with the chosen decision threshold, especially for the SVM, where increased sensitivity is accompanied by a marked rise in false positives.

**Discussion**

The results suggest that, given the available predictors, heart attack risk is largely captured by an additive and approximately linear structure on the log-odds scale. This explains why the interpretable GLM performs on par with, or better than, more flexible nonlinear models.

Although nonlinear modeling did not improve discrimination, the GAM provides additional explanatory value by illustrating how risk varies across the range of continuous predictors. This highlights that predictive performance and interpretability answer different questions and should be evaluated separately.

The strong dependence on decision thresholds further emphasizes that model usefulness cannot be judged by ROC/AUC alone. In imbalanced settings, operational priorities—such as favoring sensitivity over specificity—play a central role in determining practical applicability.

<br>

--------------------------------------------------------------------------------

<br>

# Conclusion

In this study, the binomial GLM emerged as the most suitable model, offering the best balance between predictive performance, stability, and interpretability. More complex models (SVM, GAM, ANN) did not provide meaningful gains in discrimination, indicating that increased model complexity is not necessarily beneficial for this task. Overall, the findings underscore the importance of aligning model choice and evaluation criteria with the intended risk prediction objective rather than model sophistication alone.

**Future work could be:**

- evaluate probability calibration under the original class distribution
- explore alternative imbalance handling (e.g., class weighting or SMOTE)
- validate results on clinically verified data to assess generalization beyond self-reported survey responses.

**Learnings**

- Important to tune the model to the use case and not just improve accuracy, in our application case it is more important to correctly predicte the positive heart attacks, even if it is at the cost of some false positives.

- The limited interpretability of complex models complicates effective parameter tuning.

<br>

--------------------------------------------------------------------------------


<br>

# Use of generative AI

We used ChatGPT for different supports: Explain code e.g. from labs or also for suggested code from ChatGPT itself. This works very well. For the generation of code it is more difficult, we think it is crutial to have a clear concept in mind what exactly you want to model or code and then use GenAI instead of saying here is the data, I need to use this model, give me the code. This usually doesn't work so well. Also it is important to really understand the suggested code to check its plausibility. Sometimes with a follow-up question suddenly a different approach is suggested.
<br>


<details>
<summary><b>Show Session Info</b></summary>
```{r,Sesion Info}
sessionInfo()
```
</details>